---
title: "R Notebook"
output: html_notebook
---


```{r}
set.seed(32)
#Load Libraries
library(rjags)
library(dplyr)
library(ggplot2)
library(corrplot)
library(tidyr)
```

```{r}
#Read csv
adm_data <- read.csv("admission_data.csv", header=T)
adm_data
```

#Exploratory Data Analysis

## Data Understanding

```{r}
#Structure of the dataset
str(adm_data)
```

```{r}
#Summary of the dataset
summary(adm_data)
```

## Check for Unwanted Values

```{r}
#Check for duplicate data
sum(duplicated(adm_data))
```

No duplicates.

```{r}
#Check for NA values
colSums(is.na(adm_data))
```

No NA values were found.

## Univariate Analysis

```{r}
#Transform the data
df_scaled <- as.data.frame(scale(adm_data))
```

We transform the data to improve readability.

```{r}
boxplot(df_scaled, 
        main = "Boxplots for All Variables", 
        col = rainbow(ncol(adm_data)), # Different colors for each boxplot
        las = 2,                # Rotate x-axis labels
        outline = TRUE)
```
We can see from the analysis that there are only a few outliars, but not enough to do Robust Scaling.

# Multivariate Analysis

For the multivariate analysis, we will be using a correlation plot to determine the relationship between the weights to each other and the weights to the response variable (Chance of Admit)

```{r}
cor_data <- as.data.frame(as.table(cor(adm_data)))
cor_data$Freq = round(cor_data$Freq, 2)
```

```{r}
ggplot(cor_data, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(low="red", mid = "white", high = "blue", midpoint=0) +
  geom_text(aes(label=Freq), color = "black")+
  theme_minimal()
  
  
```

> From the correlation plot, we can deduce that all 7 covariates have significance towards the response variable (Chance.To.Admit).

# Data Preprocessing

```{r}
#Initialize Data
x <- subset(adm_data, select = -Chance.of.Admit)
y <- adm_data$Chance.of.Admit
n <- nrow(adm_data)
col <- ncol(x)
```

## Data Splitting

We split the data into train and test data to accurately predict the MSE (Mean Squared Error) and overall accuracy of the model.

```{r}

#Split size
idx_train <- sample(1:n, 0.8*n)

#Split the Data
x_train = x[idx_train, ]
x_test = x[-idx_train, ]
y_train = y[idx_train]
y_test = y[-idx_train]
```

## Scaling the Data

We scale the data to help the model better find convergence.


```{r}
#Scale the data to help with convergence
x_train_scaled <- scale(x_train)
x_test_scaled <- scale(x_test)
```

#Modelling

For the model, we will use JAGS implemented in R as the basis.

```{r}
model_str <- "
  model{
    for(i in 1:N){
      Y[i] ~ dbeta(phi*mu[i], phi * (1-mu[i]))
      logit(mu[i]) <- beta_0 + inprod(beta[1:col], x[i, 1:col])
      log_lik[i] <- logdensity.beta(Y[i], mu[i] * phi, (1 - mu[i]) * phi)
    }
    
    #Priors
    beta_0 ~ dnorm(0, 1)
    
    for(i in 1:col){
      beta[i] ~ dnorm(0, 1)
    }
    phi ~ dgamma(0.001, 0.001)
  }
"
```

```{r}
#Initialize the data list
data_list <- list(
  x = x_train_scaled,
  Y = y_train,
  N = length(idx_train),
  col = col
)
```

```{r}
#Initial Values For Each Variable
inits <- function() {
  beta_inits <- rnorm(ncol(x), 0, 1)
  phi_init <- rgamma(1,2,1)
  list(
    beta_0 = rnorm(1, 0, 1),
    beta = beta_inits,
    phi = max(phi_init, 0.01)
  )
}
```

```{r}
#Create the Model
model_beta <- jags.model(
  textConnection(model_str),
  data = data_list,
  inits = inits,
  n.chains = 5
)
```

## Burn-in Process

The burn-in process is done to prevent the model to falsely diagnose the first few iterations (the learning process of the model) as divergency.

```{r}
update(model_beta, 1e3) #1000 iterations
```

## Sampling

Sampling is done to take examples from the model to be able to interpret it.

```{r}
samples <- coda.samples(
  model = model_beta,
  variable.names = c("beta_0", "beta", "phi", "log_lik"),
  n.iter = 1e4 #10000 iterations
)
```

#General Evaluation

## Summary of the Sample

```{r}
#Select only the beta variables
samples_new <- as.mcmc.list(samples[, grep("beta\\[", colnames(samples[[1]]))])
```


```{r}
#Summary of the samples
summary(samples_new)
```

```{r}
par(mar=c(2,2,2,2), mfrow=c(3,2))
plot(samples_new)
```

From this plot,we can deduce that the trace plots of each variable seems covergent. However, to further confirm this, we can do convergence diagnostics.

# Convergence Diagnostics

## Geweke Diagnostics
Geweke Diagnostics reveal the convergence z-score for each chain's parameters.

### Diagnostics

```{r}
for (i in 1:5){
  cat("Chain", i)
  print(geweke.diag(samples_new[[i]]))
}
```


### Plot

```{r}
par(mar=c(2,2,2,2))
for (i in 1:5){
  geweke.plot(samples_new[[i]])
}
```

## Gelman-Rubin Diagnostics

### Diagnostics

```{r}
gelman.diag(samples_new)
```

### Plot

```{r}
par(mar=c(2,2,2,2))
gelman.plot(samples_new)
```
## Raftery-Lewis Diagnostics
```{r}
raftery.diag(samples_new)

```
```{r}
??coda
```


## Effective Sample Size

```{r}
effectiveSize(samples_new)
```

The effective sample size indicates that the sample's phi and beta_0 has low autocorrelation, as the ESS for phi and beta_0 is close to the total iterations (25.000), but the other values is adequate for reliable inference but shows room for optimization.

## AutoCorrelation

```{r}
zero_lag_max <- 0
for(i in 1:5){
  par(mar=c(2,2,2,2))
  acf_result <- acf(samples_new[[i]], plot=F)
  lags <- acf_result$lag
  acf_values <- acf_result$acf
  zero_lag_index <- which.min(abs(acf_values))
  zero_lag <- lags[zero_lag_index]
  cat("Autocorr is 0 for chain", i, "at lag", zero_lag, "\n") 
  zero_lag_max = ifelse(zero_lag > zero_lag_max, zero_lag, zero_lag_max)
}

```

```{r}
par(mar=c(2,2,2,2))
autocorr.plot(samples_new)
```
# Model Comparison
```{r}
log_lik_matrix <- as.matrix(samples[, grep("log_lik", colnames(as.matrix(samples)))])
```

```{r}
lppd <- sum(log(colMeans(exp(log_lik_matrix))))  # Log pointwise predictive density
p_waic <- sum(apply(log_lik_matrix, 2, var))  # Effective number of parameters
cat("p_WAIC:", p_waic, "\n")
waic <- -2 * (lppd - p_waic)             # WAIC

cat("WAIC:", waic, "\n")
```

```{r}
library(loo)
log_lik_array <- array(log_lik_matrix, dim = c(nrow(log_lik_matrix), n))

# Perform LOO-CV
loo_result <- loo(log_lik_array)

# Print the result
print(loo_result)
```

# Predictions

```{r}
#Separate samples
combined_samples <- as.matrix(samples_new)

beta_0_samples <- as.matrix(samples[, "beta_0"])
beta_samples <- as.matrix(combined_samples[, grep("beta\\[", colnames(combined_samples))])
phi_samples <- as.matrix(samples[, "phi"])


n_samples <- nrow(as.matrix(samples_new))  # Number of posterior samples
n_test <- nrow(x_test)  # Number of test observations
```

```{r}
# Preallocate matrix for predicted mu
x_test_scaled <- as.matrix(x_test_scaled)
pred_mu <- matrix(NA, nrow = nrow(beta_samples), ncol = n_test)

# Compute predicted mu for each test observation

for (i in 1:nrow(beta_samples)) {
  pred_mu[i, ] <- plogis(beta_0_samples[i] + x_test_scaled %*% beta_samples[i, ])
}
```

```{r}
# Generate predictions for Y (Beta distribution)
pred_y <- matrix(NA, nrow = n_samples, ncol = n_test)
for (i in 1:n_samples) {
  for (j in 1:n_test) {
    pred_y[i, j] <- rbeta(1, phi_samples[i] * pred_mu[i, j], phi_samples[i] * (1 - pred_mu[i, j]))
  }
}
```

## Frequentist Metrics

### Model MSE (Mean Squared Error)

```{r}
mse <- mean((apply(pred_y, 2, mean) - y_test)^2)
print(mse)
```

### Baseline MSE

```{r}
baseline_mse <- mean((y_test - mean(y_train))^2)
baseline_mse
```

```{r}
pred_mean <- apply(pred_y, 2, mean)
pred_df <- data.frame(
  True = y_test,
  Predicted = pred_mean,
  Residual = y_test - pred_mean
)

pred_df
```

```{r}
ggplot() +
  geom_density(data = pred_df, aes(x = Predicted), color = "orange", fill = "orange", alpha = 0.3) +
  geom_density(data = pred_df, aes(x = True), color = "blue", fill = "blue", alpha = 0.3) +
  labs(title = "Density of True vs Predicted",
       x = "Values",
       y = "Density",
       fill = "Legend") +
  theme_minimal()
```

```{r}
#Range of Error
ggplot() +
 geom_density(data = pred_df, aes(x = Residual), color = "violet", fill = "maroon", alpha = 0.3) +
  labs(
    title = "Distribution of Residuals",
    x = "Residuals",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),  # Centered title
    axis.title = element_text(size = 12),  # Axis labels
    axis.text = element_text(size = 10)  # Axis ticks
  )
```

```{r}
library(tidyr)

pred_long <- pivot_longer(pred_df, cols = c(True, Predicted), names_to = "Type", values_to = "Value")

# Line plot for True and Predicted values
ggplot(pred_long, aes(x = seq_along(Value), y = Value, color = Type)) +
  geom_line() +
  labs(
    title = "True and Predicted Values Comparison",
    x = "Index",
    y = "Value"
  ) +
  scale_color_manual(values = c("True" = "blue", "Predicted" = "red")) +
  theme_minimal()

```

### Posterior Predicted P-Value (PPP-value)

```{r}
observed_mean <- mean(y_test)
predicted_mean <- apply(pred_y, 2, mean)

ppp_value <- mean(predicted_mean >= observed_mean)
print(paste("Posterior Predictive p-value:", ppp_value))
```

